\SW: II. New Gravitational Lens Candidates from the CFHTLS Discovered through Citizen Science

### 2015-05-09 doc/sw-cfhtls.tex
Text edits on new sample #203

In \Tref{tab:stats}, we give overall statistics of the systems
detected at \StageOne and \StageTwo. We give the total number of
detections of the known lens candidates, known lenses and the new lens
candidates at each stage.  We also give the recovery fractions for the
known samples and fraction of total detection for the new samples at
each stage. We find that all of the confirmed lenses found
at \StageOne are also recovered at \StageTwo whereas a small fraction of
the known lens candidates are rejected at \StageTwo. From the visually
inspected known lens sample at \StageTwo (P$> 0.3$), the experts gave a
grade of 1.3 and above to about 92~per~cent (33 out of 36) of the systems. Overall,
the sample of new \sw lens candidates increased the sample of total
\cfhtls lens candidates by over 50~per~cent.

@@

\begin{table}
\begin{center}
\caption{ \label{tab:stats}
Statistics of detections in \sw }
\begin{tabular}{l l l l l}
\hline
   &  \multicolumn{2}{c} {\StageOne}  & \multicolumn{2}{c}{\StageTwo} \\
      & KC  &  KL  & KC & KL \\
\hline
\hline
Number  & 128 & 34 & 107  & 34  \\
Percentage  & & & & \\
of recovery & 29 & 58 & 25 & 58  \\
P$_{\rm thresh}$ & 0.95 & 0.95 & 0.3 & 0.3 \\
{\bf Expert Grade (G)} & $-$ & $-$ & 1.3 & 1.3 \\
\hline
   &  \multicolumn{2}{c} {\StageTwo}  &   & \\
      & NC  &  AC  &  & \\
\hline
\hline
Number  & {\bf 59} & 141 &  & \\
Percentage & & & & \\
of detection & 13 & 33 &  & \\
P$_{\rm thresh}$ & 0.3 & 0.3 &  & \\
{\bf Expert Grade (G)} & 1.3 & 1.3 & 1.3 & 1.3 \\

\hline
\end{tabular}
\end{center}
{KC}-- Known lens candidates \\
{KL}-- Known lenses \\
{NC}-- New lens candidates  \\
{AC}-- All (known and new) lens candidates  \\
P$_{\rm thresh}$ -- systems with Posterior probability $P$ above this threshold are selected \\
Note: For KC and KL, Percentages are with respect to the known
population whereas for NC and AC, percentages are with respect to the
total population of lens candidates. \\
\end{table}

@@

We give basic information about the final sample of {\bf 59} new lens
candidates (medium-high probability) found by \sw in
Table~\ref{tab:swcands}. We report the candidates with a \sw ID and Name
of the lens system. We give their positions (Ra, Dec), photometric
redshift ($z_{phot}$), $i$ band magnitude of the lensing galaxy,
averaged grade G from the lens experts, zoo ID (identifier used in
TALK), $P$ value at \StageTwo and a visual categorization of the type of
lensed images and the lensing galaxy in the ``Comments'' column in
Table~\ref{tab:swcands}. Whenever available the lens properties are
taken from the \cfhtls photometric catalog \citep{Coupon2009} otherwise
for the lens galaxy positions, the reported values are measured
manually. The visual categorization of the lens type is only suggestive
and the explanation of the notations in the Comments column is given at
the bottom of the table.

We show images of our new sample in \Fref{fig:lc} which is arranged
first in the decreasing order of their grades and then increasing order
of their positions. As the first lens search was a blind search with no
preselection of candidates from any algorithm, we find various types of
lenses, as expected from such a search. The final sample consists of
both galaxy and group-scale lens candidates. There are detections of
elongated arcs and some interesting point-like quasar lensed images.
Most of them are brighter in bluer $g$ band but some candidates brighter
in the redder $i$ band are also found. Since robotic lens searches often
look for blue colored lensed features, they are very much likely to miss
such interesting lens candidates. We did not find any examples of
exotic lens candidates from the visually inspected $P>0.3$ sample. There
may be some interesting candidates lurking in TALK but these will have
to be sifted through.

The new \sw lens sample presented here shows the advantages of having
citizen scientists find lenses through visual inspection. An algorithm,
by definition will find objects that adhere to a selection criteria that
uses either geometry or flux information from an image. On the other
hand, citizen scientists can do amazing amount of interpolation or
extrapolation over the basic selection criteria provided to them.  For
example, the lower blue arc in SW7 is split by a small red galaxy.  An
algorithm typically fails to detect such arcs because the arc is broken
into smaller arclets which then falls below the minimum length or area
allowed for an arc to be detected typically. Human brains have no
problem in interpolating over the broken blue arc over the red galaxy
and understand that it is a single long arc. The system {\bf SW20} has
point-like lensed images which cannot be detected by arc finding
algorithm whereas the ring finding algorithm may miss this because of
atypical color and structure of the galaxy. Detection of red arcs, for
example, as seen in the SW38 candidate show how the volunteers
extrapolate on the color parameter because the training sample mostly
has blue arcs since this is the typical expectation.

The power of citizen scientists also lies in the high dynamic range that
allows us to find systems which vary from short (thick) to long (thin)
arcs, from highly compact to low surface brightness images, from round
and point-like to elongated and curved images, from blue to red, from
regular to exotic kinds of lenses while keeping the false positive rate
low compared to algorithms. Discovery of this large sample of completely
new candidates missed by some of these algorithms demonstrates that the
\sw system is functioning well and the self-taught citizen scientists
have managed to beat the robots in some sense.

>>>>

In \Tref{tab:stats}, we give overall statistics of the systems
detected at \StageOne and \StageTwo. We give the total number of
detections of the known lens candidates, known lenses and the new lens
candidates at each stage.  We also give the recovery fractions for the
known samples and fraction of total detection for the new samples at
each stage. We find that all of the confirmed lenses found
at \StageOne are also recovered at \StageTwo; a small fraction of
the known lens candidates are rejected at \StageTwo. From the visually
inspected known lens sample at \StageTwo (P$> 0.3$), the experts gave a
grade of 1.3 and above to about 92~per~cent (33 out of 36) of the systems. Overall,
the sample of new \sw lens candidates increased the sample of total
\cfhtls lens candidates by over 50~per~cent. However, we find that 42\% of known
lenses were missed at \StageOne: we return to the discussion of these false
negatives below.

@@

\begin{table}
\begin{center}
\caption{ \label{tab:stats}
Statistics of detections in \sw }
\begin{tabular}{l l l l l}
\hline
   &  \multicolumn{2}{c} {\StageOne}  & \multicolumn{2}{c}{\StageTwo} \\
      & KC  &  KL  & KC & KL \\
\hline
\hline
Number  & 128 & 34 & 107  & 34  \\
Fraction  & & & & \\
of total recovered & 29\% & 58\% & 25\% & 58\%  \\
$P_{\rm thresh}$ & 0.95 & 0.95 & 0.3 & 0.3 \\
{\bf Expert Grade (G)} & $-$ & $-$ & 1.3 & 1.3 \\
\hline
   &  \multicolumn{2}{c} {\StageTwo}  &   & \\
      & NC  &  AC  &  & \\
\hline
\hline
Number  & {\bf 59} & 141 &  & \\
Fraction & & & & \\
of detections & 13\% & 33\% &  & \\
$P_{\rm thresh}$ & 0.3 & 0.3 &  & \\
{\bf Expert Grade (G)} & 1.3 & 1.3 & 1.3 & 1.3 \\

\hline
\end{tabular}
\end{center}
{KC}-- Known lens candidates \\
{KL}-- Known lenses \\
{NC}-- New lens candidates  \\
{AC}-- All (known and new) lens candidates  \\
$P_{\rm thresh}$ -- systems with Posterior probability~$P$ above this threshold are selected \\
Note: For KC and KL, Percentages are with respect to the known
population whereas for NC and AC, percentages are with respect to the
total population of lens candidates. \\
\end{table}

@@

We give basic information about the final sample of {\bf 59} new medium grade
lens candidates found by \sw in
\Tref{tab:swcands}. We report the candidates with a \sw ID and Name
of the lens system. We give their positions (RA, Dec), photometric
redshift ($z_{phot}$), $i$ band magnitude of the lensing galaxy,
averaged grade~$G$ from the lens experts, zoo ID (identifier used in
TALK), $P$ value at \StageTwo and a visual categorization of the type of
lensed images and the lensing galaxy in the ``Comments'' column in
\Tref{tab:swcands}. Whenever available the lens properties are
taken from the \cfhtls photometric catalog \citep{Coupon2009}; otherwise,
for the lens galaxy positions, the reported values were measured
manually. The visual categorization of the lens type is only suggestive
and the explanation of the notations in the Comments column is given at
the bottom of the table.

We show images of our new sample in \Fref{fig:lc}. The panels are arranged
first in the descending order of their grades, and within each grade, in
ascending order
of RA. As the first lens search was a blind search with no
preselection of candidates from any algorithm, we find various types of
lenses, as expected from such a search. The final sample consists of
both galaxy and group-scale lens candidates. There are detections of
elongated arcs and some interesting point-like quasar lensed images.
Most of them are brighter in the bluer $g$ band, but some candidates brighter
in the redder $i$ band are also found. Since the robotic lens searches
focused on the blue lensed features, they are likely to miss
such interesting lens candidates. We did not find any examples of
exotic lens candidates from the visually inspected $P>0.3$ sample.

The new \sw lens sample presented here illustrates some of
the advantages of having
citizen scientists find lenses through visual inspection. An algorithm,
by definition will find objects that adhere to a selection criteria that
uses either geometry or flux information from an image. On the other
hand, citizen scientists can interpolate over or
extrapolate beyond the basic selection criteria provided to them.  For
example, the lower blue arc in SW7 is split by a small red galaxy.  An
algorithm typically fails to detect such arcs because the arc is broken
into smaller arclets which then falls below the minimum length or area
allowed for an arc to be detected. Human inspectors have no
problem in interpolating over the broken blue arc over the red galaxy,
understanding that it is a single long arc. The system SW20 has
point-like lensed images which cannot be detected by arc finding
algorithm, whereas the ring finding algorithm may have missed this because of
the atypical color and structure of the galaxy. Detection of red arcs, for
example, as seen in the SW38 candidate, show how the volunteers
extrapolate on the color parameter: the training sample contains
predominantly blue arcs, because the source colors were drawn from realistic
observed distributions.

The power of citizen scientists also lies in the high dynamic range that
allows us to find systems which vary from short (thick) to long (thin)
arcs, from highly compact to low surface brightness images, from round
and point-like to elongated and curved images, from blue to red, from
regular to exotic kinds of lenses while keeping the false positive rate
low compared to algorithms. Discovery of this large sample of completely
new candidates missed by some of these algorithms demonstrates that the
\sw system is functioning well, the self-taught citizen scientists
reaching parts of discovery space that the robots did not.
